---
title: 'Lab 9: Mapping of Traits'
author: "Adam Davis and Elizabeth McBride"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, results = FALSE, warning = FALSE, message = FALSE)
```

***
Please read all materials in labs and run all code! TIP: Don't just run the example code in the labs. **Inspect the results you get**, and see how they relate to the text around them.

Some of today's lab was adapted from:
https://cran.r-project.org/web/packages/sf/vignettes/sf1.html
***

##### Clean environment
```{r}
rm(list=ls())
```

##### Load packages
```{r}
library(tidyverse)
```



#### Load and combine person / household data

```{r}
PersonData <- read_rds('Data/PersonData_111A.Rds')
HHData <- read_rds('Data/HHData_111A.Rds')
hh_bgDensity <- read_rds('Data/hh_bgDensity.Rds')

personHHData <- left_join(PersonData, HHData) %>%
  left_join(hh_bgDensity)
```


#### Spatial data

You will be installing a new package today: the `sf` package. This is a package for working with spatial data.

```{r}
install.packages("sf")
library(sf)
```


You will be adding new data to your project today: a shapefile for making maps!

1. Go to Gauchospace and download the newest data
2. Add this set of files to a new folder in your project named "Spatial". The data will already be within a folder of this name, so you can 
3. Once you have added it, continue with the instructions below



All of the `sf` functions for dealing with spatial data start with `st_`. To read our new shapefile into R, we will use `st_read()`
```{r}
counties_shp_dat <- st_read("Spatial/CA_counties.shp")

counties_shp <- counties_shp_dat %>% 
  rename(CTFIP = GEOID) %>%
  mutate(CTFIP = CTFIP %>% as.character%>%as.numeric)



st_write(counties_shp, "Spatial/counties.shp", delete_dsn=TRUE)
```

```{r}
counties_shp <- st_read("Spatial/counties.shp")
```

First quick plot. The unique colors correspond to the unique names of each county:
```{r}
plot(counties_shp["NAME"])
```



### Attach some traits using `left_join`

A great benefit of `sf` is its ability to work within the `tidyverse`

We will now be taking some of the person and household characteristics and converting them to a format where they can be . We will be aggregating characteristics to the county level so we can map the traits of our CHTS respondents county-by-county

1. 

Variables where you want the mean:
```{r}
prhh_aggreg <- personHHData %>% 
  mutate(Sum_VMT = DriveAlone_Dist + Driveothers_Dist +Passenger_Dist) %>%
  select(hhid, pnum, CTFIP, County, bg_group,Sum_Trips, Sum_PMT, Sum_VMT, bornUSA, starts_with("pers")) %>%
  group_by(CTFIP) %>%
  summarise_at(vars(-hhid, -pnum, -County, -bg_group, -CTFIP), funs(mean,n()))

#divide by the number of households in each. Actually can just do the mean for each county
  
```

For the variables where it's a 1 or 0 (like bornUSA, or any of the variables that start with "pers"), this number is the percentage of people who answered "yes" (aka 1) to . I checked whether excluding the DK/RF answers (which are coded as 8 or 9 so could theoretically mess with the means) was necessary or changed the means and it made a minimal difference so we won't worry about it for now.

3. Now, you can simply join the two datsets together using `left_join` like we have done in the past
```{r}
county_prhh_shp <- counties_shp %>% left_join(prhh_aggreg)
```


4. First map of traits
```{r}
mapview(county_prhh_shp, 
        zcol = "bornUSA", legend=TRUE, 
        label= county_prhh_shp$NAME, popup = popupTable())
```
