---
title: "Geog 111A Lab 3"
author: "Elizabeth McBride"
date: "Fall 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, results = FALSE, warning = FALSE, message = FALSE)
```

## Introduction

This script introduces the `%>%` operator for multi-step processes. It will walk you through the `tidyverse` functions `mutate()`, `summarize()`, and `group_by()`. It will also include an exploration of the data we will be using from the California Household Travel Survey (CHTS).

**NOTE:** Don't forget to use the built-in "Help" if you don't know what a function does (e.g. type `?mutate` to learn more about it). You can also find more information on the `dplyr` cheat-sheet linked at the beginning of the assignment.

**Please be sure to read everything! Also be sure to run every line of code in this assignment on your own to follow along!**

***
<span style="color:red">For submission: One script with the code asked for in the questions marked in red. Please make sure you have your **first name initial and full last name** in your document name. </span>

***


To use functions from a package, **you need to load the package library**:
```{r, eval = TRUE}
library(tidyverse)
```
You must **load the library at the beginning of every new R session.** This gives you access to all the functions in the packages listed in your console when you run the line.

#### Load person data
Same as last week, we'll be using a simplifed version of the dataset `peopleFull`, and name it `people`.
```{r, eval = TRUE}
peopleFull <- read_rds('Data/PersonData_111A.Rds')
```
The line above reads in the file `PersonData_111A.Rds` dataset under the name `peopleFull`

#### Grab only the columns you want
```{r, eval = TRUE}
people <- select(peopleFull, hhid:persOthr,DriverLic:Employed,Student,workday,Walk_trips:Allothers_trips,Sum_Trips,Sum_PMT)
```
**REMINDER:** This line selects columns by name from the dataset `peopleFull`, and creates a new dataset with those columns that you name `people`. You should have 25 variables and 94,901 observations in the newly-created `people` file. Check the `Environment` box in RStudio to ensure this.

##### <span style="color:red">Q1. *Before beginning calculations, you will examine some of the variables you will be using.* Open your Data Dictionary in a spreadsheet software and look at the PersonData_111A sheet. In commented sentences in your script, answer the following. (A) What is `Sum_PMT` for each person? (B) what is `Sum_Trips`? (C) What is the difference between the columns ending in `_Dist` and the columns ending in `_trips` (3) </span>


## `%>%` -- multi-step processes with the pipe operator

Because each function in `dplyr` is relatively simple, it is often necessary to call multiple functions in sequence to perform an operation.

As you may have noticed, the input dataset is the first input to all `dplyr` functions. Because of this, `dplyr` recommends the `%>%` operator (which comes from the `tidyverse` to chain functions together in sequence. 

`%>%` works by entering the **object on the left into a function as the first argument**. This means that instead of reading functions inside-out, you can read them left-to-right.

So `mean(people$Age)` is the same as `people$Age %>% mean()`. This can be repeated as many times in one operation as you want!
```{r}
round(mean(people$Age), digits = 1) # first compute mean, then round
people$Age %>% mean() %>% round(digits = 1) # same, just (maybe) easier to read
```
You first take the mean of `people$Age`, then round it. Note how the first function needs to be read "inside-out" to interpret it, while the second one is read left-right.

**PRO TIP** The keyboard shortcut for `%>%` is `ctrl + shift + m` on a PC or `cmd + shift + m` on a Mac.

#### So with `dplyr` functions, if you want to select a subset of rows and columns:
```{r}
people %>% select(hhid:Age) %>% filter(Age < 10)
```

##### <span style="color:red">Q2. Examine the subset you just selected. (A) What are the names of all the columns grabbed in this selection process? (B) What did the `filter()` function do? (C) How many observations (rows) total were included in the selection (don't forget to include the first 10 rows printed in the console in this number)?  </span>


##### <span style="color:red">Q3. Show me how to write the line of code `people %>% select(hhid:Age) %>% filter(Age < 10)` in the "inside-out" way (i.e. the way we've been writing code up until now). (3) </span>

#### Another benefit of this is that it lets RStudio autocomplete column names!
To try this, enter this in the console: `people %>% select(Ag`  
then hit `<Tab>`

## `mutate()` -- calculate new columns
It’s often useful to **create new columns** from your dataset. One method to do this in base R using `$` notation is shown in earlier labs. In the `tidyverse`, you would use `mutate()`.

Look up `mutate()` in the help system, and look at the "Arguments" section. The second argument is `...`, and it is descibed as "Name-value pairs of expressions." This means the "name" goes on the left side of `=`, and the "value" goes on the right side of `=`. Basically, it means we can name our columns that we calculate the values of within the same `mutate()` expression.

We will start by creating a column with each respondent's average trip length.

**NOTE on multi-line commands** Commands can span multiple lines, as long as the lines are connected by a simple operator (`+-*/`), a `%%` operator (e.g. `%>%` `%in%`), or are within parentheses. You can use `Ctrl+Enter`/`Command+Enter` on any line within a command and it will run the command as a whole. In the following code, we will put new lines between the two separate function calls `select()` and `mutate()`. **As you type the code below for yourself, notice that Rstudio even automatically indents when you press "enter" within an unfinished command.**
```{r}
people %>% 
  select(hhid, pnum, Sum_PMT, Sum_Trips) %>% # Select a small subset of only columns we need
  mutate(mean_TrpLength = Sum_PMT / Sum_Trips) # Divides miles traveled by total trips
```
<span style="color:red">  Before moving on, **make sure you understand how the calculation above worked.** Look at the console output from the code above. Compare the rowwise values of `Sum_PMT` and `Sum_Trips` to the `mean_TrpLength` column you created. (not a question for points). </span>


#### You can use variables created in a `mutate()` call within the same call:
```{r}
people %>% 
  select(hhid, pnum, Sum_PMT, Sum_Trips) %>%
  mutate(mean_TrpLength = Sum_PMT / Sum_Trips,
         half_mean = mean_TrpLength / 2) 
```
Notice that within the same `mutate()` function call, we created `mean_TrpLength`, then used `mean_TrpLength` to calculate values for another column we named `half_mean`. **Before moving on, examine your console output and compare the values of `mean_TrpLength` to those of `half_mean` and  make sure you understand the code you just ran.**  If you do not see `mean_TrpLength` column, check for this in the console: `... and 1 more variable`. If you see that, just make your Rstudio window or the console panel wider so more text can fit, then run again.

##### <span style="color:red">Q4. Show me the code to create a new variable where you divide the respondents' ages in half. Please use the pipe operator `%>%`. You can name the variable whatever you wish. (2)</span>

## `summarize()` / `summarise()` -- calculate aggregate statistics from data
To learn about the **overall** characteristics of your dataset, use the `summarize()` (or `summarise()`) command, which **condenses your data into a single row**.

<span style="color:red">Run the following, and **make sure you understand what each command did** before moving on to the next.</span>

#### Average age of respondents:
```{r}
people %>% summarize(meanAge = mean(Age))
```

#### Average age, number of trips, distance, and average trip length:
```{r}
people %>% summarize(meanAge = mean(Age),
                     meanPMT = mean(Sum_PMT),
                     meanTrips = mean(Sum_Trips),
                     meanLength = meanPMT / meanTrips)
```

#### Maximum PMT and trips:
```{r}
people %>% summarize(maxPMT = max(Sum_PMT),
                     maxTrips = max(Sum_Trips))
```
Notice how you can list multiple arguments separated with a comma when you want more than one summarization.

## Using summary functions in `mutate()`:
Summary functions can also be used to create new variables using `mutate()`. For example, below we use `mean()` within `mutate()` to subtract the **overall mean** from every person's PMT.

#### so to get difference betweem a person's PMT and the mean PMT in the sample:
```{r}
people %>% select(hhid,pnum, Sum_PMT) %>% 
  mutate(PMTdif = Sum_PMT - mean(Sum_PMT))
```

#### just to make sure, this value should have mean 0:
```{r}
people %>% select(hhid,pnum,Sum_PMT) %>% 
  mutate(PMTdif = Sum_PMT - mean(Sum_PMT)) %>%
  summarise(mean(PMTdif))
```
Which it basically does!

##### <span style="color:red">Q5. Show me the code to return the difference between a person's total Number of Trips by all modes and the mean Number of Trips by all modes. (4) </span>


## `group_by()` -- Defining groups for a dataset
Many datasets contain multiple observations that are related to each other. For instance, you may have multiple observations from a certain town or from a certain group of interest. Really, any categorical variable can be used for grouping if it provides an interesting comparison.

In this dataset, variables can potentially be grouped by household id, various social, demographic, and economic variables, weekday, and access to modes.


### Counting potential groups
`dplyr` makes it easy to find out how many people have any value of a variable or set of variables.

```{r}
people %>% count(Male)
people %>% count(Employed,Student) # How many people have each combination of employment and student status
```


##### <span style="color:red">Q6. (A) How many respondents identify as male, and how many do *not* identify as male? (B) How many respondents are employed AND a student? *(Remember to check the Data Dictionary for descriptions of the variables and their values.)* (2) </span>


`dplyr` interprets grouped data differently from ungrouped data, but first you have to define grouping, using the `group_by()` function.


```{r}
people %>% 
  select(hhid, pnum, Male, Age) %>% 
  group_by(Male)
```
Notice `Groups:   Male [2]` written at the top of the console output? It names the grouping variable and has the total # of groups in brackets. In this case, one variable (Male) and 2 groups.

### Grouping using logical operators
```{r}
people %>% group_by(Sum_PMT < 10 | Sum_PMT > 20) %>% summarise(How_Many = n())
```
This creates two groups from the Person Miles Traveled (PMT) variable. The `TRUE` group contains people who traveled less than 10 miles on their travel diary day, OR traveled more than 20 miles on their diary day. The `FALSE` group contains everyone else (i.e., those who traveled between 10 miles and 20 miles). 

##### <span style="color:red">Q7. Use `group_by()` and `summarise()` to find out how many people are between age 25 and 65 (including those that are 25 and 65). How many people are outside of that age range? Show your code and include your numeric answers as comments  (3) </span>

### You can also stack **groups across multiple variables**

```{r}
people %>%
  select(hhid, pnum, Male, Age, Employed) %>% 
  group_by(Male, Employed) # two variables, six groups
```
Why did grouping two binary variables produce six groups? We will find out in the following section.


## Group summaries -- using `group_by()` and `summarize()` together
One major use of grouping is to **investigate the differences between groups.**

`summarize()` is used the same way for groups as it is used for the whole dataset, but instead of providing a one-line answer, it provides one line for each group.


**NOTE:** summarizing a grouped dataset removes the lowest layer of grouping, so:  
`dataset %>% group_by(A,B) %>% summarize(count=n())`  
would produce a summary dataset still grouped by `A`, with each unique combo of variables `A` and `B` occupying one row.


Back to the question from earlier (Why did grouping two binary variables produce six groups?) Turns out there are some NA's in the Employed variable:
```{r}
people %>% group_by(Male, Employed) %>% summarise(How_Many = n())
```


The 6 groups:

* not male & not employed
* not male & employed
* not male & NA (not yet 16)
* male & not employed
* male & employed
* male & NA (not yet 16)


##### <span style="color:red">Q8. (A) Look up the help page for `n()`, and describe what `n()` returns in one sentence (your own words). (B) Based on this information, in one sentence, describe what the `How_Many` column we created contains. (2) </span>


##### <span style="color:red">Q9. Use `group_by()` and `summarise()` to find out how many people identify as hispanic and white, and how many identify as hispanic and non-white. Show the code you used and give your numeric answers in a comment. *(Hint: find out which columns you need in the data dictionary)* (5) </span>


#### Comparing (non-statistically) trip counts between work/nonwork days for employed people:
```{r}
people %>% filter(Employed == 1) %>% 
  group_by(workday) %>%
  summarise(nPeople=n(), meanTrips = mean(Sum_Trips))
```
Above, we first used `filter()` to grab only survey respondents who are employed. 

##### <span style="color:red"> Q10. (A) Add an argument to the `summarise()` function above to additionally calculate the mean PMT for workdays versus non-workdays. Show your code. (B) Interpret your results in a couple commented sentences. How do workers' travel behaviors differ on workdays vs. non-workdays? (3) <span>


#### Combining What You Learned: 

##### <span style="color:red">Q11. Do people in their prime work years (between ages 25 and 65) travel more miles than and make more trips than people of other ages? Or are their travel patterns indistinguishable from the travel patterns of other age groups? Please include your code on your answer sheet. *Group by this age category and compare the mean of Person Miles Traveled for this age group with the mean of the rest of the sample. Do the same thing for Number of Trips. Include everyone, not just employed people. In a couple commented sentences, interpret the results.* (5) </span>

