---
title: 'Geog 111A Lab 2: Important Concepts in R and the Tidyverse'
author: "Written by Elizabeth C. McBride and Adam W. Davis"
date: "Fall 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

This lab will walk you through some more important concepts for learning R. **Please be sure to read everything! Also be sure to run every line of code in this assignment on your own to follow along!**

***
<span style="color:red">For submission: 
One script with the code asked for in the questions marked in red. Please make sure you have your **first name initial and full last name** in your document name. </span>

***

## Creating an Rstudio Project (only if using RStudio desktop)

Projects simplify the code process by keeping all the data files and scripts in a "local" directory. You can move the folder containing the project, and all contained scripts will still be able to run. It removes the need for telling R exactly where every data file you want to access "lives" in your computer. **From now on, when you do your labs, you will save any new scripts or data to the project folder you will create below:**

1. Click File > New Project > New Directory
2. In the "Project Type" window, click "New Project"
3. In the "Create New Project" window under "Directory Name", name your project (e.g. `yourSurname_GEOG111A_Project.Rproj`)
4. Under "Create project as subdirectory of", click "Browse..." and navigate to the folder location where you would like to store your 111A lab materials for the rest of the quarter
5. Leave everything else the same, click "Create Project". This should open a new RStudio window.
6. In the top right corner of the Rstudio window, you should now see the name of the project file. This confirms that you are now working within the 111A project. 


More information on projects can be found [here] (<https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects>)

### Adding a new script in your project folder
When you create the project, a folder will be created that contains your `.Rproj` file. This folder is where you will store all datasets and scripts you want to use for a project. You will now add your first script to your project folder.

1. Create a new script by File > New File > R Script
2. Save the file with whatever name you prefer (e.g. `yourlastname_Lab3.R`) in the folder that contains your R project.

### Organizing your project folder

1. Navigate to your project folder in File Explorer (PC) / Finder (Mac)
2. Create a new folder named "Data"
3. Create another folder named "Outputs"


### Adding data to your project

1. From Gauchospace, download the dataset(s) for today's lab.
2. Add them to the `/Data` folder you created

The files added to your `/Data` folder will never be directly edited. They should stay the same as the day you downloaded them. Any edited versions will be saved to the "Output" folder. This is to ensure that you maintain a "clean" copy of the original files.


## Recap

### Assigning Names to Objects

We create objects using `<-`, with the name we want to give the object on the left side, and the value(s) we want to give the object on the right. 

Like so:
```{r}
myobject <- 5
```

Once you run that line, R will know that if you type `myobject` you are referring to the number `5`. Throughout your code, it will treat `myobject` **exactly the same** as if you typed the number `5`:

```{r}
myobject^2 # R will treat this line exactly the same as the next line
5^2
```

HOWEVER, if you run another line of code later where you redefine your object (assign the same name new values), any code run afterwards referencing that object will use the most recent assignment values.
```{r}
myobject <- 55

myobject^2
```

### Vectors (DO NOT SKIP, MAKE SURE YOU RUN THIS CODE)
The `c()` function "combines its arguments to form a vector" (from `c()` function help page). Run the following:
```{r}
range15 <- c(11,12,13,14,15)
range15 # run this line to look at your new object in the console
```
You have just created a numeric vector object! It is named `range15`, and contains the integers from 11 to 15. You should see it in your `Environment` panel. 

Vectors can consist of any data type. Below, you will create a vector of strings:
```{r}
listOfModes <- c('Auto','Bike','Foot','Bus','Rail')
listOfModes
```


## Data Frames

One particularly useful data structure in R is the `data.frame`, which stores data in an array of named columns (rows can also be named, but you probably don't want to). All the elements of a given column must be of the same type, but a `data.frame` can store multiple types of columns.

For instance, this creates a data frame with one numeric column and one text column:
```{r}
df1 <- data.frame(r15 = range15, mode = listOfModes, stringsAsFactors = FALSE)
df1 # run this line to view your new data frame in the Console
```
Note how we were able to name the columns using `=` within the `data.frame` function. You will see the names of the columns at the top of each one in your `Console` pane. The argument `stringsAsFactors = FALSE` makes sure that `data.frame()` function properly reads in `listOfModes` as a column of character values (don't worry about this for now!).  

You can also view data frames in a spreadsheet-like format using the `View()` command:
```{r}
View(df1)
```
ALSO: Clicking `df1` in the `Environment` pane (upper right corner of Rstudio) has the same effect as running `View(df1)`.

In the `Environment` pane, another way to examine your dataset is to click the blue triangle to the left of `df1`. Try it now!

### Elements of a data frame can be accessed using column names

```{r}
df1$r15 # this will grab all items in the column called r15
```
The `$` allows you to select a column of the data frame by its name. See `?Extract` for more information on this and other functions for extracting specific values from your data.


### New columns can be created and assigned to the table in much the same way

This creates a new column named `double_r15` that is equal to two times the value of `r15` in each row:
```{r}
df1$double_r15 <- df1$r15 * 2
View(df1)
```
Notice how when you run `View(df1)` there is now a third column named `double_r15`? We just created that! And notice how the values in `double_r15` are the values of column `r15` multiplied by two? 


##### <span style="color:red">Q1. Create another new variable within `df1` named `r15_sq` that consists of the squared values of `r15`. Include your code. (2) </span>

## Summarizing and Visualizing Data

R's basic functionality includes a number of very useful functions for summarizing data. Let's try a few, using a built-in R dataset called `mtcars`.

We'll start by looking up this built-in dataset so we can get a little more information about it:
```{r}
?mtcars
```
Notice in the `Help` window it gives you the name of the column and then the description of the column. Next, we will store the dataset as an object and view it as a table:
```{r}
mtcars <- mtcars #this just saves the dataset by name in our environment
View(mtcars)
```


### Calculate the mean and median vehicle weights (1000 lbs):
```{r}
mean(mtcars$wt, na.rm = TRUE)
median(mtcars$wt, na.rm = TRUE)
```

Just to show you an example of why we put the `na.rm` argument into the function above, sometimes there will be NA's in your data. Some functions, like `mean()` or `median()`, are sensitive to NA values:
```{r}
testing_NA <- c(1,2,3,NA,5)
mean(testing_NA)
mean(testing_NA, na.rm = TRUE)
```
Notice the different outputs of the two mean calculations. An `NA` is not a number, it is just a blank space. If you think about it, it makes sense why R does not know how to include that in the calculation of the mean.

##### <span style="color:red">Q2. (A) Look up the "help" page for `NA` (case-sensitive). In your own words, what would an `NA` value in the data mean? (B) Now Look up the "help" page for `mean` or `median`. In your own words, what does `na.rm` mean? (C) Why would you use the argument `na.rm = TRUE` in your mean or median calculation? (3) </span>


##### <span style="color:red">Q3. Show me the code to calculate the mean and median of the `mpg` variable in the `mtcars` dataset. (2) </span>



## Loading External Packages

### This command will install a number of packages that will be useful this week and for the rest of the quarter:

If you are working on Rstudio Desktop and you haven't already done this step on the **computer you are currently using**, start by installing the `tidyverse`. This will install a number of packages on the computer that will be useful this week and for the rest of the quarter:
```{r}
install.packages('tidyverse')
```
If you do not have admin priveleges, you may be asked to create a personal library the first time you run this.


### In order to use a package you have installed, you need to load the package library:
```{r}
library(tidyverse)
```
This gives you access to all of the functions in the packages listed in your console when you run the line.
**NOTE:** You may see a warning about `readr` being built with an earlier version of R. Do not worry about this, everything will still work.



## Packages and Functions

#### Functions

**Functions** let you do things in R without having to write complicated code from scratch. Earlier, we used the `mean()` function. This allows us to calculate the mean of something without having to carefully write out the entire calculation from scratch, like so: `(1+5+8+8+10+22+17)/7`

Some functions are built in to base R (like `mean()`), and some you get from external packages you install. You can also create your own.

#### Packages

To use functions from a package, **you need to load the package library**:
```{r}
library(tidyverse)
```
You must **load the library at the beginning of every new R session.** This gives you access to all the functions in the packages listed in your console when you run the line.

**NOTE:** You may see some warnings about packages being built with an earlier version of R. Do not worry about this, everything will still work. You may see some other warnings about objects being "masked" from other packages. That is fine too.


## Introducing our data for the quarter

**CHTS Data Dictionary**

1. Download the Data Dictionary from Gauchospace (`DataDictionary.xlsx`)
2. Using a spreadsheet software, open the Data Dictionary. 
3. Take some time to study the first three sheets: "Data File Guide", "PersonData_111A", and "HHData_111A". Make sure to examine the variables, descriptions, and values.

You will definitely want to keep this Data Dictionary handy throughout the quarter. It has all kinds of valuable information about the datasets we are using. 

##### <span style="color:red">Q4. (A) In a couple commented sentences, describe the types of variables contained in `PersonData_111A`. What information does it provide about respondents? (B) Do the same for `HHData_111A`. (C) Notice that `hhid` is present in both datasets. How might this be useful to us? (6) </span>



## Read and Write Files

When you want to work with data you already have on your computer, you will need to `read` it into R and give it a name:
```{r}
peopleFull <- read_rds(path = "Data/PersonData_111A.Rds")
View(peopleFull) # run this to examine the person dataset you just read in
```
One convenient feature of using a Project is that it does not need you to write that whole directory! Only subdirectories (like `Data/`).

In this class, we'll mostly be using this R-specific data format, `.Rds` (R dataset) since it will load more quickly and consistently on everyone's computer.

**IMPORTANT NOTE:** When you read in a file to R, any changes you make will **NOT** be changing your original dataset file. The original stays exactly the same. You are essentially working with a copy of that dataset that you have loaded into R. You can save changes using a `write` function. For now, you do not need to use a `write` function for anything, so we will get into detail about this when it comes up in future labs.

## Tidyverse 
In this section, you will learn about some of the functions from the `tidyverse` we will use the most this quarter.

The `tidyverse` is NOT a package itself. Rather, it is a collection of packages with various functionalities. Combined, these packages create a powerful set of tools for data science. Starting now, the code and sequence of this lab is adapted from the `dplyr` vignette "Introduction to dplyr" found [here](<https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html>)

**Tips:**

* Packages in the `tidyverse` have well crafted documentation, including [cheat sheets](<https://www.rstudio.com/resources/cheatsheets/>) (which I like to have open at all times):


* There are also detailed code vignettes to help you learn the packages on your own. The vignettes for `dplyr` available [here](<https://cran.r-project.org/web/packages/dplyr/index.html>)


### Select -- select and rename columns
Very often, you'll be working with datasets that contain way more columns than you need. In this case, it's very cumbersome to carry around all 50 columns in the people dataset

#### First, display all the column names, so you know what you're working with:
```{r}
names(peopleFull)
```
Notice they have the same names as the variables from the DataDictionary?


#### It's easy to select a single colum in the basic R:
```{r}
ageCol <- peopleFull$Age 
```
Rstudio adds this to your `Environment` (top left) as a Large Numeric, under Values

#### But grabbing multiple columns is a MAJOR pain. You can use ranges of numbers, if you figure them out:
```{r}
peopleFull[,1:10] 
```
This selects the first 10 columns, from `hhid` to `persPacIsl`

#### But otherwise, you have to name columns in a character vector, eg:
```{r}
peopleFull[,c('hhid','pnum','Age')]
```

#### The function `select` from the `dplyr` package provides a number of easy solutions to this problem. You can name columns individually:
```{r}
select(.data = peopleFull, hhid, pnum, Age)
```
**Note:** the first entry is the dataset name. The next 3 are the names of the columns you want.


#### Or ask for all columns in a certain range using `:`
```{r}
select(peopleFull, hhid:persOthr, Sum_Trips, Sum_PMT) 
```
This code selects all cols between `hhid` and `persOthr`, plus `Sum_Trips` and `Sum_PMT`. 

#### Or even grab columns based on the format of their name!
```{r}
select(peopleFull, hhid:Age, starts_with('pers'), starts_with('Sum'))
```
This addition selects columns that start with the string "pers", and columns that start with "Sum"

#### It's easy to rename variables too using the equal sign in the selection:
```{r}
select(peopleFull, personAge = Age) # selects ONLY the Age column, renames it personAge
rename(peopleFull, personAge = Age) # keeps all columns, renames Age to personAge
```
You can rename columns when you select the ones you want. If you want to keep all the columns and just rename some of them, you can use the function `rename`.

#### or select specific columns to **exclude**, using a `-` sign:
```{r}
select(peopleFull, -persDKrace, -persRFrace) 
```

##### <span style="color:red">Q5. Provide the code you would use to `select` all columns from `peopleFull` EXCEPT the column indicating if they were born in the USA? Check your DataDictionary if you need clues. (1) </span>

We'll be using a simplifed version of the dataset we will name `people` for the rest of the script. Created below:
```{r}
people <- select(peopleFull, hhid:Age, starts_with('pers'), starts_with('Sum'),-persDKrace, -persRFrace)
```
 Check your `people` table by pressing the blue drop-down arrow to the left of its name in the "Environment" window of RStudio. It should have the following 13 columns:
`hhid`, `pnum`, `Male`, `Age`, `persHisp`, `persWhite`, `persAfricanAm`, `persNativeAm`, `persAsian`, `persPacIsl`, `persOthr`, `Sum_Trips`, and `Sum_PMT`.

### Filter -- select rows

#### First, a reminder of the logical operators:
```{r}
10 == 10 # TRUE
10 >= 9  # TRUE
10 < 3   # FALSE
10 != 10 # FALSE
!TRUE    # FALSE
```
For quick descriptions of these and other operators, check this site: http://www.statmethods.net/management/operators.html


#### "AND" and "OR"
You can use `&` for AND, and `|` for OR. This allows you to include multiple logical operators in one TRUE/FALSE statement. `&` is used if you want to know when **both conditions are true at the same time**. `|` is used if you want **one condition, OR the other condition, OR both** to be true at the same time. 

#### You'll often want to perform analysis on a specific subset of the observations in your data.
We already learned how to do that in base R. Here's how to extract only **males over the age of 75** in base R:
```{r}
people[people$Age > 75 & people$Male == 1,] 
```
NOTE: the `,<blank>` (comma followed by nothing) means grab all columns.

#### The `dplyr` syntax is simpler, use the `filter` function:
```{r}
filter(.data = people, Age > 75, Male == 1)
```

#### or the equivalent
```{r}
filter(.data = people, Age > 75 & Male == 1)
```

#### For boolean variables / variables where all values are either 0 or 1, you can just do this:
```{r}
filter(people, Age > 75 & Male)
```
Because...  
1 == `TRUE`  
0 == `FALSE`

The console should show that all of the above formulations select the same 1,947 people!

##### <span style="color:red">Q6. (A) What does a 0 mean for the `Male` variable? (B) Show me the code to select only people who are 25 years old and African American? (2) </span>

#### ALSO NOTE: these functions have not modified the original object `people`!
To store the changes you've made, you have to save the function's output to a variable, eg:
```{r}
nonagenarians <- filter(people, Age >= 90)
```


#### Filter operations using OR are somewhat more restricted, but still straightforward:
```{r}
filter(people, persAfricanAm == 1 | persNativeAm ==1)
```
#### or the equivalent (because they are Boolean variables)
```{r}
filter(people, persAfricanAm | persNativeAm)
```
Both of these select the same 9,402 people

### Arrange -- sort data
Sometimes, you'll want to rearrange your data based one or more variables

#### In base R, this is achieved with bracket notation and the `order()` function:
```{r}
ppl_age <- select(people, hhid, pnum, Age, Sum_Trips) #making a smaller dataset to simplify the visualization
sampToSort <- sample_n(ppl_age,10) # create a random sample of 10 people look at
sampToSort$Age # run this to look at ages of people in sample
order(sampToSort$Age) # returns rank order of ages. 1st entry is the index of the lowest observation
sampToSort # run this to view the original
sampToSort[order(sampToSort$Age),] # order applied to rows. Compare to the original (run line above this one)
```
When you run those last two lines, you should see that the table has been sorted based on `Age`. 


#### `dplyr` makes this simpler by taking one step out of the process:
```{r}
arrange(sampToSort,Age)
```

#### to sort in the opposite direction, use the `desc()` function:
```{r}
arrange(sampToSort,desc(Age))
```



### Distinct -- get unique values
The process for grabbing unique observations is basically the same between R basic and `dplyr`, but is MUCH faster in `dplyr`.

#### here's how to grab the unique household id's:
```{r}
distinct(people, hhid) # there are 36,925 unique households in this version of the dataset
```

##### <span style="color:red">Q7. Show me the code to grab the unique person numbers. This will basically give us the number of distinct household sizes present in the dataset. If you dont know which variable is the person number, use your Data Dictionary. (2) </span>